---
title: "Red Wine Data Exploration and Summary"
output: html_document
---
#### **by Eric Spitler**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning=FALSE, message=FALSE}
# load ggplot and psych
library(ggplot2)
library(psych)
```

```{r}
redWine <- read.csv('wineQualityReds.csv')
```

### **Introduction**

This report explores a dataset containing 11 variables on the chemical properties of 1,599 red wines. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent). The 13th variable listed in the dataset dimensions below is an index column.

### **Univariate Plots Section**

**Dataset dimensions**
```{r}
dim(redWine)
```

**Dataset overview**
```{r}
str(redWine)
```

The dataset contains 13 variables with 1599 observations. The first variable, 
'X', is an index. Variables 2-12 measure the chemical properties of the red 
wines. Variable 13 is the quality rating of the wine as assessed by experts. 

**Summary Statistics of Variables**
```{r}
summary(redWine)
```

##### **Exploration Goal**
I want to understand how the chemical properties of red wine relate to quality
rating. Which properties correlate most with the highest and lowest rated wines?
Are there any interesting relationships between the various chemical properties 
of the wines?

##### **Histogram of Quality**
```{r}
# PLOT 1 - quality histogram
ggplot(aes(x = as.factor(quality)), data = redWine) +
 geom_bar()
```

All of the wines in the dataset have a rating between 3 and 8, with most rated 5 or 6.

##### **Histograms of all chemical properties of the wines**
```{r}
# PLOT 2 - fixed acidity histogram
ggplot(aes(x = fixed.acidity), data = redWine) +
  geom_histogram(bins = 11)
```

Fixed acidity is right skewed with a mean that appears to be around 8 or 9.

```{r}
# PLOT 3 - volatile acidity histogram
ggplot(aes(x = volatile.acidity), data = redWine) +
  geom_histogram(bins = 11)
```

Volatile acidity is strongly right skewed. A few wines have a volatile acidity
level roughly three times the mean.

```{r}
# PLOT 4 - citric acid histogram
ggplot(aes(x = citric.acid), data = redWine) +
  geom_histogram(bins = 11)
```

Citric acid is tightly grouped in the bottom 50% of the range. The distribution
has a skinny tail at the top of the range. 

```{r}
# PLOT 5 - residual sugar histogram
ggplot(aes(x = residual.sugar), data = redWine) +
  geom_histogram(bins = 11)
```

Residual sugar has a very skinny long tail. A few wines in the dataset have a 
significantly higher residual sugar level than nearly all of the rest of the 
wines.

```{r}
# PLOT 6 - chlorides histogram
ggplot(aes(x = chlorides), data = redWine) +
  geom_histogram(bins = 11)
```

The chloride levels for the majority of the wines is tightly distributed with
a long skinny tail on the right. 

```{r}
# PLOT 7 - free sulfur dioxide histogram
ggplot(aes(x = free.sulfur.dioxide), data = redWine) +
  geom_histogram(bins = 11)
```

The free sulfur dioxide distribution is right skewed with a fat long tail on the 
right. 

```{r}
# PLOT 8 - total sulfur dioxide histogram
ggplot(aes(x = total.sulfur.dioxide), data = redWine) +
  geom_histogram(bins = 11)
```

Very few wines have a total sulfur dioxide level that is significantly higher
than the rest of the wines. This is another right skewed distribution.

```{r}
# PLOT 9 - free sulfur dioxide histogram
ggplot(aes(x = density), data = redWine) +
  geom_histogram(bins = 11)
```

Free sulfur dioxide is roughlt normally distributed. There is a wide range of
free sulfur dioxide levels across the wines.

```{r}
# PLOT 10 - pH histogram
ggplot(aes(x = pH), data = redWine) +
  geom_histogram(bins = 11)
```

pH also is roughly normally distributed. 

```{r}
# PLOT 11 - sulphates histogram
ggplot(aes(x = sulphates), data = redWine) +
  geom_histogram(bins = 11)
```

Sulphate levels are very right skewed with a long skinny tail. A few wines have 
sulphate levels 3-4x higher than the mean.

```{r}
# PLOT 12 - alcohol histogram
ggplot(aes(x = alcohol), data = redWine) +
  geom_histogram(bins = 11)
```

Alcohol is right skewed but there is a fat right tail, so there is a number of
wines throughout the distribution of alcohol levels.

```{r}
# subset the data by quality rating so we can get summary stats of every
# chemical property for each quality rating, and whatever else we want
q3 <- subset(redWine, quality == 3)
q4 <- subset(redWine, quality == 4)
q5 <- subset(redWine, quality == 5)
q6 <- subset(redWine, quality == 6)
q7 <- subset(redWine, quality == 7)
q8 <- subset(redWine, quality == 8)
```

I want to compare the chemical properties for wines with the worst and best
quality ratings and see if this helps me identify which properties to examine
more closely.

**Summary statistics for each variable for wines with a quality rating of 3**

```{r}
summary(q3)
```

**Summary statistics for each variable for wines with a quality rating of 8**
```{r}
summary(q8)
```

# Univariate Analysis

### What is the structure of your dataset?
There are 1599 red wines in the data set, with 12 features (quality, fixed
acidity, volatile acidity, citric acid, residual sugar, chlorides, 
free sulfur dioxide, total sulfur dioxide, and density). The quality rating is 
an ordered factor, and while it is a numerical variable, it is really a 
categorical variable.

  * Most wines have a quality rating of 5 or 6. None of the wines were rated as
  low as 0, 1, or 2. None of them were rated as high as 9 or 10.
  
  * The mean value for each chemical property is as follows
      + fixed acidity - 8.32
      + volatile acidity - 0.5278
      + citric acid - 0.271
      + residual sugar - 2.539
      + chlorides - 0.08747
      + free sulfur dioxide - 15.87
      + total sulfur dioxide - 46.47
      + density - 0.9967
      + pH - 3.311
      + sulphates - 0.6581
      + alcohol - 10.42
      + quality - 5.636
      
  * Density and pH are both normally distributed, and the distributions of the 
    other chemical properties have varying degrees of right-skewness.

### What is/are the main feature(s) of interest in your dataset?
The main feature of the dataset is quality. My interest in examining the data
is to understand the chemical properties of the wines that correlate with a 
wine receiving a specific expert quality rating. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I want to further explore the relationships between quality and volatile
acidity, citric acid, chlorides, pH, sulphates, and alcohol. This is because 
for each of these chemical properties there is a linear trend from the mean
value for wines with a quality rating of 3, to the dataset mean, to the mean 
value for wines with a quality rating of 8. My hypothesis is that these
chemical properties will help explain the quality ratings of the wines.

### Did you create any new variables from existing variables in the dataset?
I did not create any new variables. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Since the dataset was already very tidy, I didn't need to change anything.

# Bivariate Plots Section

Looking at scatter plots for each chemical property I want examine more closely
vs. quality lets me quickly see how each property is distributed for each 
quality rating.

```{r}
# PLOT 13 - scatter plot - quality vs. volatile acidity
ggplot(aes(x = quality, y= volatile.acidity), data=redWine) + 
  geom_jitter(alpha = 1/10)

```

The distribution of volatile acidity is the lowest and tightest for the highest
rated wines.

```{r}
# PLOT 14 - scatter plot - quality vs. citric acid
ggplot(aes(x = quality, y= citric.acid), data=redWine) + 
  geom_jitter(alpha = 1/10)
```

Citric acid doesn't tell a strong story.

```{r}
# PLOT 15 - scatter plot - quality vs. chlorides
ggplot(aes(x = quality, y= chlorides), data=redWine) + 
  geom_jitter(alpha = 1/10)
```

The distribution of chlorides for the highest rated wines is the tighest and
lowest of all. 

```{r}
# PLOT 16 - scatter plot - quality vs. pH
ggplot(aes(x = quality, y= pH), data=redWine) + 
  geom_jitter(alpha = 1/10)
```

All quality ratings have a wide range of pH values.

```{r}
# PLOT 17 - scatter plot - quality vs. sulphates
ggplot(aes(x = quality, y= sulphates), data=redWine) + 
  geom_jitter(alpha = 1/10)
```

Wines rated 3 and 4 have some of the lowest means sulphates. The 
distribution of sulphates in the highest rated wines are tightly grouped.

```{r}
# PLOT 18 - scatter plot - quality vs. alcohol 
ggplot(aes(x = quality, y= alcohol), data=redWine) + 
  geom_jitter(alpha = 1/10)
```

While each rating has relatively wide distribution of alcohol levels, in 
general, as alcohol goes up, quality rating goes up.

```{r}
# create a subset containing the chemical properties I want to look closer at
chem_props <- redWine[, c(3,4,6,10,11,12,13)]
```

Let's look at a correlation plot of the chemical properties and quality. The 
plot will also give a brief overview of any other interesting relationships
between any of the variables.

```{r}
# Plot 19 - correlation plot for chem props and quality
pairs.panels(chem_props)
```

The strongest correlation is between alcohol and quality, besides the 
correlation between pH and citric acid, which makes intuitive sense, since 
more acid means a lower pH.

Because of the lack of strong relationships in the above correlation plot,
I'm curious to look at all the variables

```{r}
# Plot 20 - correlation plot for all variables 
pairs.panels(redWine)
```

I was wondering if quality and residual sugar had any relationship. Since the
reviewers are experts, not casual wine drinkers, sweetness doesn't play a
factor here. The correlation coefficient is 0.01!

Leaving out relationships between pH and other acidity properties, the strongest
correlations among chemical properties exist between density and fixed acidity,
and between sulphates and chlorides.

Quality, volatile acidity, and alcohol share relatively high correlations, 
so I was curious to see them graphed together.

```{r}
# Plot 21 - scatter plot of alcohol/volatile acidity vs. quality
ggplot(aes(x = quality, y= alcohol/volatile.acidity), data=redWine) + 
 geom_jitter(alpha = 1/10)
```

The lowest rated wines all have very low ratios of alcohol to volatile acidity.

I'm curious to see if the proprtion of sulfur dioxide that is 'free' is any 
different across quality ratings. 

```{r}
# Plot 22 - scatter plot of free sulfur dioxide/total sulfer dioxide vs. quality
ggplot(aes(x = quality, y= free.sulfur.dioxide/total.sulfur.dioxide), 
       data=redWine) + geom_jitter(alpha = 1/10)
```

Volatile acidity negatively correlates with quality and fixed acidity correlates
positively, albeit weakly. I'm curious if the ratio of volatile to fixed acidity
relates interestingly with quality

```{r}
# Plot 23 - scatter plot of volatile acidity/fixed acidity vs. quality
ggplot(aes(x = quality, y= volatile.acidity/fixed.acidity), 
       data=redWine) + geom_jitter(alpha = 1/10)
```

With one exception, the highest rated wines have a low ratio of volatile to 
fixed acidity. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
The strongest (although not really that strong) correlations between quaility and
chemical properties exists with alcohol (r = 0.48), volatile acidity (r = -0.39)
, sulphates (r = 0.25), and citric acid (r = 0.23). Of the chemical properties,
only alcohol has what qualifies as a moderately strong relationship. Volatile 
acidity is right on the edge of having a moderately strong negative relationship
with quality. 

More than any other quality rating, the worst rated wines all shared a low 
ratio of alcohol to volatile acidity. The lowest ratio of alcohol to volatile
acidity of all belonged to wine with a quality rating of 3. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
The few strong relationships between chemical properties exist between

  * Fixed acidity and density: r = 0.67
  * Fixed acidity and citric acid: r = 0.67 (yes, both are 0.67)
    The more of a specific acid, the higher the overall acid.
  * Alcohol and density: r = -0.50
    This makes sense because water's density is 997 kg/m^3 and ethanol's density
    is 789 kg/m^3, so the higher the proportion of alcohol in the wine, the
    lower the density will be.
    
Examining the ratio of volatile acidity to fixed acidity in the wines showed
that all but one of the highest rated wines shared a low ratio. All the other
quality ratings had wide distributions of this ratio. 

### What was the strongest relationship you found?
The strongest relationship is a tie between the relationships with fixed acidity
and density and fixed acidity and citric acid. The strongest relationship with
quality, the main feature of interest is with alcohol. 

# Multivariate Plots Section

```{r}
# PLOT 24 - scatter plot - density vs. fixed acidity by quality
dfa <- ggplot(redWine, aes(x=density, y=fixed.acidity, color=factor(quality))) +
  geom_point(alpha = 1/9)
dfa +labs(title = 'Density vs. Fixed Acidity by Quality', 
          x= 'Density', y= 'Fixed Acidity')
```

A moderately strong linear relationship is visible here, but we can see that
there are no strong groupings of quality in the data. The dominance of 5 and 6 
quality rated wines is also apparent. 

```{r}
# PLOT 25 - scatter plot - sulphates vs. chlorides by quality  
dfa <- ggplot(redWine, aes(x=sulphates, y=chlorides, color=factor(quality))) +
  geom_point(alpha = 1/9)
dfa +labs(title = 'Sulphates vs. Chlorides by Quality', x= 'Density', y= 'Fixed Acidity')
```

We can clearly see that there is no linear relationship between density and 
fixed acidity, even though it looked promising in the bivariate analysis. We see
that most wine are under 1.0 for density and 0.15 for fixed acidity, with the 
larger subsets, 5 and 6 having more outliers.

```{r}
# PLOT 25 - scatter plot - alcohol vs. volatile acidity by quality  
dfa <- ggplot(redWine, aes(x=alcohol, y=volatile.acidity, 
  color=factor(quality))) +
  geom_point(alpha = 1/9)
dfa +labs(title = 'Alcohol vs. Volatile Acidity by Quality', 
          x= 'alcohol', y= 'volatile acidity')
```

Here we can see a concentration 5-rated wines in the lower-left quadrant, but
other than that there's not much of a relationship apparent.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
None of the relationships between chemical properties and quality that I 
explored showed any strong interactions. This is an interesting dynamic in the 
dataset.

### Were there any interesting or surprising interactions between features?
The linearity of the relationship between density and fixed acidity was 
surprising. There may be some underlying chemical principals that explains this,
but this was not obvious at the outset. I was surprised that were wasn't an 
interaction of chemical properties that related more strongly to quality. 

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
dfa <- ggplot(redWine, aes(x=density, y=fixed.acidity, color=factor(quality))) +
  geom_point(alpha = 1/10)
dfa +labs(title = 'Density vs. Fixed Acidity', x= 'Density', y= 'Fixed Acidity')
```

### Description One
There is a moderately strong relationship (r=0.67) between fixed acidity and 
density. However, by observing the color in the scatterplot to identify trends
in the relationship between quality and fixed acidity/density, we can see that
there isn't a strong trend.

### Plot Two
```{r echo=FALSE, Plot_Two}
aq <- ggplot(aes(x = quality, y= alcohol), data=redWine) + 
  geom_jitter(alpha = 1/10)
aq +labs(title = 'Alcohol by Quality Rating',
         x = 'Red Wine Quality Rating',
         y = 'Percent  Alcohol')
```

### Description Two
This plot clearly shows that Quality Rating increases with Alcohol. The lowest
rated wines had a tight distribution of low Alcohol percentage. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ava <- ggplot(aes(x = quality, y= alcohol/volatile.acidity), data=redWine) + 
  geom_jitter(alpha = 1/10)
ava + labs(title = 'Alcohol/Volatile Acidity by Quality Rating', 
         x = 'Red Wine Quality Rating', 
         y = 'Ratio of Alcohol to Volatile Acidity')
```

### Description Three
This plot shows that the lowest rated wines all have some of the lowest ratios
of alcohol to volatile acidity. The highest rated wines have the tighest 
distribution of alcohol/volatile acidity of any of the other quality ratings.

------

# Reflection

The red wine dataset contains 11 chemical properties and an expert quality
rating for 1599 wines. Using R made it easy to quickly acquaint myself with the
dataset and begin exploring relationship in the data. Since this dataset had one 
main variable of interest, quality, it made sense to compare quality to all the 
chemical properties. It was fairly easy to see some trends in the relationships
between quality and a few chemical properties. In examining the relationships
between the chemical properties, the strongest  relationship is between
fixed acidity and density. There is also a moderately strong relationship 
between fixed acidity and citric acid, but this is to be expected. 

In examining the data, I set out to focus my bivariate analyis on the chemical
properties that had a linear trend from the mean of the lowest rated wines to
the dataset mean to the mean of the highest rated wines. This lead me quickly
to some moderately explanatory graphs, but again, nothing that was terribly 
strong and clear. 

It was a bit frustrating looking for relationship in the data in general. There
was no 'smoking gun' or big, obvious story in the data. It would be interesting
to examine a dataset with ratings from a single expert and try to determine what
role the preferences of the expert rater plays in the relationships between
chemical properties and quality rating. It would also be interesting to compare
ratings of untrained wine drinkes to those of experts for the same wines. I have
read articles in the past that highlighted the inabilities of wine raters to 
consistently rate the same wine. This dynamic may contribute to the fact that
the chemical property with the strongest correlation to quality was alcohol, 
which doesn't seem like a property that would strongly effect the quality of the
wine within a reasonable range. 

To investigate the dataset further, I would build a linear model to understand
which variables or interactions of variables best explain the quality ratings of
the wines. A dataset with wines from more regions outside Portugal might provide
less skewed distributions for the chemical properties, which could enhance the 
predictive power of a linear model. 
